---
title: "CodeBook"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Coursera Getting and Cleaning Data
### *Smartphone Activity Recognition Project*

## Project Goals

Create an R script named "run_analysis.R" that performs the following:

1. Merge the training and the test sets to create one data set.
2. Extract only the measurements on the mean and standard deviation for each measurement.
3. Use descriptive activity names to name the activities in the data set
4. Appropriately label the data set with descriptive variable names.
5. From the data set in step 4, create a second, independent tidy data set with the average of each variable for each activity and each subject.

## Data Overview
Source data is from the UCI Machine Learning Repository (http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones) and consists of a Human Activity Recognition database built from the recordings of 30 subjects performing activities of daily living (ADL) while carrying a waist-mounted smartphone with embedded inertial sensors.

The experiments were carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, 3-axial linear acceleration and 3-axial angular velocity were captured at a constant rate of 50Hz. The experiments were video-recorded to label the data manually. The dataset is initially randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data.

## Initial Variables

The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

* tBodyAcc-XYZ
* tGravityAcc-XYZ
* tBodyAccJerk-XYZ
* tBodyGyro-XYZ
* tBodyGyroJerk-XYZ
* tBodyAccMag
* tGravityAccMag
* tBodyAccJerkMag
* tBodyGyroMag
* tBodyGyroJerkMag
* fBodyAcc-XYZ
* fBodyAccJerk-XYZ
* fBodyGyro-XYZ
* fBodyAccMag
* fBodyAccJerkMag
* fBodyGyroMag
* fBodyGyroJerkMag

The set of estimated variables that will be used in this project include: 

* mean(): Mean value
* std(): Standard deviation

Additional vectors were obtained by averaging the signals in a signal window sample. These are used on the angle() variable:

* gravityMean
* tBodyAccMean
* tBodyAccJerkMean
* tBodyGyroMean
* tBodyGyroJerkMean

## Analysis Walkthrough

```{r ,message=F}

# Load the required packages
require(dplyr)


# Gather feature labels
feature.labels <- read.table("features.txt",header=FALSE)


# Gather activity labels
activity.labels <- read.table("activity_labels.txt",header=FALSE)
# provide better column names, "activity" column name will be used later in a table join
colnames(activity.labels) <- c("activity","activity_type")


# Gather measured data for training and test
train.data <- read.table("train/X_train.txt",header=FALSE)
test.data <- read.table("test/X_test.txt",header=FALSE)


# Gather activity classification data for training and test
train.activity <- read.table("train/y_train.txt",header=FALSE)
test.activity <- read.table("test/y_test.txt",header=FALSE)


# Gather subject data
train.subject <- read.table("train/subject_train.txt",header=FALSE)
test.subject <- read.table("test/subject_test.txt",header=FALSE)


# Combine all training and test data
combined.data <- rbind(train.data,test.data)
combined.activity <- rbind(train.activity,test.activity)
combined.subject <- rbind(train.subject,test.subject)


# Make relevant column labels
colnames(combined.activity) <- "activity"
colnames(combined.subject) <- "subject"
colnames(combined.data) <- feature.labels$V2


# Combine data sets
all.data <- cbind(combined.subject,combined.activity,combined.data)


# Remove duplicate columns
all.data.trimmed <- all.data[, !duplicated(colnames(all.data))]


# Extract only the measurements on mean and standard deviation
all.data.filtered <- all.data.trimmed %>% select(subject,activity,contains("mean"),contains("std"))


# Add human readable activity labels
# Perform table join to associate activity number with activity text, then remove the activity number
all.data.labeled <- all.data.filtered %>% left_join(activity.labels,by = "activity") %>% select(-activity)


# Provide more detailed variable names and remove duplication
names(all.data.labeled) <- gsub("^t", "time", names(all.data.labeled))
names(all.data.labeled) <- gsub("^f", "frequency", names(all.data.labeled))
names(all.data.labeled) <- gsub("Acc", "Accelerometer", names(all.data.labeled))
names(all.data.labeled) <- gsub("Gyro", "Gyroscope", names(all.data.labeled))
names(all.data.labeled) <- gsub("Mag", "Magnitude", names(all.data.labeled))
names(all.data.labeled) <- gsub("BodyBody", "Body", names(all.data.labeled))

```

```{r table-data-labeled}
knitr::kable(head(all.data.labeled))
```

```{r, warning=F}
# Group the final dataset by subject and activity type, and then summarize by the mean of each variable
all.data.mean <- all.data.labeled %>% group_by(subject,activity_type) %>% summarize_all(funs(mean))
```

```{r table-data-mean}
knitr::kable(head(all.data.mean))
```

